---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gt)
library(readr)
```

## Data preparation work before generating the summary table

Import the data table.

```{r}
tbl <- read_rds(file = "data-14_2_01.rds")
```

Take a look at the structure of the table.

```{r}
tbl
```

Get the current date with the `Sys.Date()` function.

```{r}
reporting_date <- as.character(Sys.Date())
```

Get the row with the *N* values.

```{r}
n_values_row <- tbl %>% filter(category == "Age (y)")
```

Pull each of the *N* values for the required columns.

```{r}
placebo_n <- n_values_row %>% pull(placebo)
xanomeline_ld_n <- n_values_row %>% pull(xanomeline_ld)
xanomeline_hd_n <- n_values_row %>% pull(xanomeline_hd)
total_n <- n_values_row %>% pull(total)
```

Add in a footnote from an external Excel file (here, called `titles.xlsx`).

```{r}
# Get the table that contains all text for different table types
titles <- readxl::read_xlsx("data/titles.xlsx")

# Extract the footnote text for the p-value column
p_footnote <-
  titles %>%
  filter(table_number == "14-2.01" & type == "footnote" & index == 1) %>%
  pull(text1)
```


## Summary table generation with **gt**

This is the final table.

```{r}
gt_table <-
  tbl %>%
  gt() %>%
  tab_header(
    title = "Table 14.2.01",
    subtitle = "Summary of Demographic and Baseline Characteristics",
    preheader = c("Protocol: CDISCPILOT01", "Population: Intent-to-Treat")
  ) %>%
  fmt_missing(columns = everything(), missing_text = "") %>%
  fmt_number(
    columns = c(placebo, xanomeline_ld, xanomeline_hd, total),
    rows = label %in% c("Mean", "SD", "Median", "Min", "Max"),
    n_sigfig = 3
  ) %>%
  fmt_number(columns = p, decimals = 4) %>%
  fmt_percent(
    columns = ends_with("_pct"),
    decimals = 0,
    scale_values = FALSE
  ) %>%
  cols_merge_n_pct(col_n = placebo, col_pct = placebo_pct) %>%
  cols_merge_n_pct(col_n = xanomeline_ld, col_pct = xanomeline_ld_pct) %>%
  cols_merge_n_pct(col_n = xanomeline_hd, col_pct = xanomeline_hd_pct) %>%
  cols_merge_n_pct(col_n = total, col_pct = total_pct) %>%
  cols_label(
    placebo  = md(paste0("Placebo  \n(N=", placebo_n, ")")),
    xanomeline_ld = md(paste0("Xanomeline  \nLow Dose  \n(N=", xanomeline_ld_n, ")")),
    xanomeline_hd = md(paste0("Xanomeline  \nHigh Dose  \n(N=", xanomeline_hd_n, ")")),
    total = md(paste0("Total  \n(N=", total_n, ")")),
    p = "p-value[1]"
  ) %>%
  tab_footnote(footnote = p_footnote) %>%
  tab_source_note(
    source_note = paste('Program Source: 14-2.01.R     Executed: (Draft)', reporting_date)
  ) %>%
  cols_width(
    1 ~ pct(25),
    2 ~ pct(20)
  ) %>%
  tab_options(
    page.orientation = "landscape",
    page.numbering = TRUE,
    page.header.use_tbl_headings = TRUE,
    page.footer.use_tbl_notes = TRUE
  )
```

Write the **gt** table to an RTF document.

```{r}
gt_table %>% gtsave("tbl_14.2.01.rtf")
```
