---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(gt)
library(tidyverse)
```

# The *Create Table* Family of Functions

All **gt** tables begin with a call to `gt()`, where we would supply the input data table (data frame or tibble) and some basic options for creating a stub (`rowname_col`) and row groups (`groupname_col`). We can also introduce a grouped tibble (via **dplyr**’s `group_by()`) for more precise divisions of rows into row groups. The `gt_preview()` function is great for getting a nicely-formatted preview of a data table (defaulting to the first 5 rows and the last row).

### `gt()`: Create a **gt** table object

The `gt()` function creates a gt table object when provided with table data. Using this function is the first step in a typical **gt** workflow. Once we have the **gt** table object, we can perform styling transformations before rendering to a display table of various formats.

##### EXAMPLE

Create a table object using the `exibble` dataset; use the `row` and `group` columns to add a stub and row groups.

```{r}
tab_1 <- exibble %>% gt(rowname_col = "row", groupname_col = "group")

tab_1
```

---------------------------------------------------------------------

The resulting object can be used in transformations (with `tab_*()`, `fmt_*()`, `cols_*()` functions).

```{r}
tab_1 %>%
  tab_header(
    title = "Table Title",
    subtitle = "Subtitle"
  ) %>%
  fmt_number(
    columns = num,
    decimals = 2
  ) %>%
  cols_label(num = "number")
```

---------------------------------------------------------------------

### `gt_preview()`: Preview a **gt** table object

Sometimes you may want to see just a small portion of your input data. We can use `gt_preview()` in place of `gt()` to get the first x rows of data and the last y rows of data (which can be set by the `top_n` and `bottom_n` arguments).

##### EXAMPLE

Use `gtcars` to create a **gt** table preview (with only a few of its columns). You'll see the first five rows and the last row.

```{r}
gtcars %>%
  dplyr::select(mfr, model, year) %>%
  gt_preview()
```

---------------------------------------------------------------------
